On the Gravitational stability of gravito-turbulent accretion disks

### 2016-03-28 paper/model3d.tex
revised version done

\section{Basic equations}\label{basic}
We consider a 3D, self-gravitating,
viscous disk with heating and cooling. We use the shearing box framework to
study a small patch of the disk \citep{goldreich65}. The local frame
co-rotates with a fiducial point in the unperturbed disk at angular
frequency $\Omega$. The Cartesian co-ordinates $(x,y,z)$ correspond
to the radial, azimuthal and vertical directions in the global disk. 
The fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) = 0, \\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi - 2\Omega\hat{\bm{z}}\times\bm{v}\notag\\ & 
  \phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =}
  +2\Omega^2qx\hat{\bm{x}} -\Omega_z^2z\hat{\bm{z}} 
   + \frac{1}{\rho}\nabla\cdot\bm{T},\label{momentum_eq}\\
  & \frac{\p E}{\p t} + \nabla\cdot\left(E\bm{v}\right) = - P
  \nabla\cdot\bm{v} + \mathcal{H}_\mathrm{visc} - \Lambda +
  \mathcal{H}_\mathrm{ext} ,\label{energy_eq}
\end{align} 
where $\rho$ is the density field and $\bm{v} = (v_x, v_y, v_z)$ is
the velocity field. 
We assume an ideal gas so that the pressure $P$ and thermal energy
density $E$ are related by  
\begin{align}\label{ideal_eos}
  P = (\gamma-1)E = \mathcal{R}\rho T, 
\end{align}
where $\mathcal{R}$ is the gas constant and $T$ is the temperature. 
{\bf For simplicity we only refer to the adiabatic index as that
  appears in Eq. \ref{ideal_eos} for both 2D and 3D disk models 
  \citep[cf.][]{johnson03}.  %should adopt diff values in 2D and 3D
                             %for the same 
                             %disk in 2D 
%johnson: specify 3D gamma, then translate into 2D gamma because disk
%model is 2D 
}
The gas 
gravitational potential $\Phi$ is given via the Poisson equation,  
\begin{align}
 \nabla^2\Phi = 4 \pi G \rho. 
\end{align}

@@

\subsection{Steady states and cooling models}
We consider equilibrium solutions (here omitting the `eq' subscripts
for simplicity)  
\begin{align}
  \bm{v} &= -q\Omega x \hat{\bm{y}}, \\
  \rho   &= \rho(z),\\
  P      &= P(z) \equiv c_s^2(z)\rho.
\end{align} 
The equilibrium density and pressure fields are obtained by solving
the vertical momentum equation with self-gravity,
\begin{align}
  &\frac{1}{\rho}\frac{dP}{dz} +
  \Omega_z^2z + \frac{d\Phi}{dz} = 0, \label{vert_eq1}\\
 &\frac{d^2\Phi}{dz^2} = 4 \pi G \rho,\label{vert_eq2}
\end{align}
together with thermodynamic equilibrium,
\begin{align}\label{thermal_eq}
(q\Omega)^2\rho\nu + \mathcal{H}_\mathrm{ext} = \Lambda,
\end{align}
where the first term represents viscous heating. For the viscous
problem, $\nu\neq0$, and we set $\mathcal{H}_\mathrm{ext}=0$ to 
obtain a relation between viscous heating and cooling 
(e.g. Eq. \ref{alpha_beta_relation} below). However,
if we wish to neglect viscosity (and the accompanying dissipation) but include cooling, we must invoke
$\mathcal{H}_\mathrm{ext}\neq0$ to define an equilibrium state.  
To proceed further, we separately describe the two cooling models
considered in this work.

\subsubsection{Beta cooling}\label{beta_cool_model}
In our beta cooling model, the energy loss per unit volume is specified  
as an explicit function of the thermodynamic variables. 
A prototypical example is 
\begin{align}\label{beta_cool}
  \Lambda(\rho, T) =
  %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c},
\end{align}
{\bf where $T_\mathrm{irr}$ is a reference temperature field, and
recall}  $\tcool = \beta\Omega^{-1}$ is 
the cooling timescale with $\beta$ a constant input parameter. 
Physically, $\tirr$ {\bf may} the floor temperature set by, 
for example, stellar or background irradiation.

>>>>

\section{Basic equations}\label{basic}
We consider a 3D, self-gravitating,
viscous disk with heating and cooling. We use the shearing box framework to
study a small patch of the disk \citep{goldreich65}. The local frame
co-rotates with a fiducial point in the unperturbed disk at angular
frequency $\Omega$. The Cartesian co-ordinates $(x,y,z)$ correspond
to the radial, azimuthal and vertical directions in the global disk. 
The fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) = 0, \\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi - 2\Omega\hat{\bm{z}}\times\bm{v}\notag\\ & 
  \phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =}
  +2\Omega^2qx\hat{\bm{x}} -\Omega_z^2z\hat{\bm{z}} 
   + \frac{1}{\rho}\nabla\cdot\bm{T},\label{momentum_eq}\\
  & \frac{\p E}{\p t} + \nabla\cdot\left(E\bm{v}\right) = - P
  \nabla\cdot\bm{v} + \mathcal{H}_\mathrm{visc} - \Lambda +
  \mathcal{H}_\mathrm{ext} ,\label{energy_eq}
\end{align} 
where $\rho$ is the density field and $\bm{v} = (v_x, v_y, v_z)$ is
the velocity field. 
We assume an ideal gas so that the pressure $P$ and thermal energy
density $E$ are related by  
\begin{align}\label{ideal_eos}
  P = (\gamma-1)E = \mathcal{R}\rho T, 
\end{align}
where $\mathcal{R}$ is the gas constant and $T$ is the temperature. 
{\bf For simplicity we refer to the adiabatic index $\gamma$ as that
  in Eq. \ref{ideal_eos} for both 2D and 3D disk models 
  \citep[cf.][]{johnson03}.  %should adopt diff values in 2D and 3D
                             %for the same 
                             %disk in 2D 
%johnson: specify 3D gamma, then translate into 2D gamma because disk
%model is 2D 
}
The gas 
gravitational potential $\Phi$ is given via the Poisson equation,  
\begin{align}
 \nabla^2\Phi = 4 \pi G \rho. 
\end{align}

@@

\subsection{Steady states and cooling models}\label{steady_state}
We consider equilibrium solutions (here omitting the `eq' subscripts
for simplicity)  
\begin{align}
  \bm{v} &= -q\Omega x \hat{\bm{y}}, \\
  \rho   &= \rho(z),\\
  P      &= P(z) \equiv c_s^2(z)\rho.
\end{align} 
The equilibrium density and pressure fields are obtained by solving
the vertical momentum equation with self-gravity,
\begin{align}
  &\frac{1}{\rho}\frac{dP}{dz} +
  \Omega_z^2z + \frac{d\Phi}{dz} = 0, \label{vert_eq1}\\
 &\frac{d^2\Phi}{dz^2} = 4 \pi G \rho,\label{vert_eq2}
\end{align}
together with thermodynamic equilibrium,
\begin{align}\label{thermal_eq}
(q\Omega)^2\rho\nu + \mathcal{H}_\mathrm{ext} = \Lambda,
\end{align}
where the first term represents viscous heating. For the viscous
problem, $\nu\neq0$, and we set $\mathcal{H}_\mathrm{ext}=0$ to 
obtain a relation between viscous heating and cooling 
(e.g. Eq. \ref{alpha_beta_relation} below). However,
if we wish to neglect viscosity (and the accompanying dissipation) but include cooling, we must invoke
$\mathcal{H}_\mathrm{ext}\neq0$ to define an equilibrium state.  
To proceed further, we separately describe the two cooling models
considered in this work.

\subsubsection{Beta cooling}\label{beta_cool_model}
In our beta cooling model, the energy loss per unit volume is specified  
as an explicit function of the thermodynamic variables. 
A prototypical example is 
\begin{align}\label{beta_cool}
  \Lambda(\rho, T) =
  %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c},
\end{align}
{\bf where $T_\mathrm{irr}$ is a reference temperature field, and
recall}  $\tcool = \beta\Omega^{-1}$ is 
the cooling timescale with $\beta$ a constant input parameter. 
Physically, $\tirr$ {\bf may be} the floor temperature set by, 
for example, stellar or background irradiation.

### 2016-03-28 paper/paper3d.tex
revised version done

\begin{abstract}
  Low mass, self-gravitating accretion disks admit quasi-steady,
  `gravito-turbulent' states in which cooling balances turbulent viscous 
  heating. However, numerical simulations show that gravito-turbulence
  cannot be sustained beyond dynamical timescales when the cooling
  rate or corresponding turbulent viscosity is too large. The result is disk fragmentation.         
  We motivate and quantify an interpretation of disk 
  fragmentation as the inability to maintain gravito-turbulence due to  
  formal secondary instabilities driven by:  
\begin{inparaenum}[1)] 
\item 
  cooling, which reduces pressure support; and/or
\item 
  viscosity, which reduces rotational support. 
\end{inparaenum}
%We analyze the gravitational stability of accretion disks 
%including these non-ideal effects explicitly.    
%We derive a general dispersion relation for linear waves in a 
%self-gravitating, viscous, non-adiabatic disk with internal heating,
%external irradiation, and cooling. 
We analyze the {\bf axisymmetric} gravitational stability of viscous, non-adiabatic
accretion disks with internal heating, external irradiation, and
cooling {\bf in the shearing box approximation}.   
We consider parameterized cooling functions in {\bf 2D} and {\bf 3D}
disks, %two and three dimensions,  
as well as   
radiative diffusion in 3D. We show that 
generally there is no critical cooling rate/viscosity below which 
the disk is formally stable, although interesting limits appear 
for unstable modes with lengthscales on the order of the disk thickness.   
%when we examine the growth 
%of modes of order the disk scale height. 
We apply this new linear theory to 
models of protoplanetary disks subject to
gravito-turbulence modeled as an effective 
viscosity, and cooling regulated by dust opacity. 
We find that viscosity renders the disk beyond $\sim 60$AU dynamically
unstable on radial %this is for Mdot = 1d-6  
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. 
We suggest turbulent stresses can play an active role in realistic disk 
fragmentation by removing rotational stabilization against 
self-gravity, and that the observed transition in behavior from 
gravito-turbulent to fragmenting may derive from  
instability of the gravito-turbulent state itself.
\end{abstract}

@@

{\bf 
\subsection{Instability of the gravito-turbulent state}\label{gi_of_gi}
  In this work, we formally treat the gravito-turbulent
  disk described above as an equilibrium state, to which we apply    
  standard linear stability analysis. Thus by `perturbations' we mean   
  deviations away from this gravito-turbulent 
  basic state\footnote{This should not be confused with 
    fluctuations with respect to the 
    laminar disk that maintain the underlying
    gravito-turbulence.}. 
  We assume instabilities defined this way  
  signifies disk fragmentation, since numerical simulations  
  show that the non-existence of steady gravito-turbulence implies 
  fragmentation.   
}

Indeed, we find that the region where a gravito-turbulent PPD is
dynamically unstable agrees with the current consensus on  where, and
the conditions under which, a PPD is expected to fragment based on
Eq. \ref{frag_cond}, with typical critical values cited in the
literature \citep[e.g. $\beta_c\sim
  3,\,\alpha_c\sim0.1$,][]{gammie01,rice05}.

>>>>

\begin{abstract}
  Low mass, self-gravitating accretion disks admit quasi-steady,
  `gravito-turbulent' states in which cooling balances turbulent viscous 
  heating. However, numerical simulations show that gravito-turbulence
  cannot be sustained beyond dynamical timescales when the cooling
  rate or corresponding turbulent viscosity is too large. The result is disk fragmentation.         
  We motivate and quantify an interpretation of disk 
  fragmentation as the inability to maintain gravito-turbulence due to  
  formal secondary instabilities driven by:  
\begin{inparaenum}[1)] 
\item 
  cooling, which reduces pressure support; and/or
\item 
  viscosity, which reduces rotational support. 
\end{inparaenum}
%We analyze the gravitational stability of accretion disks 
%including these non-ideal effects explicitly.    
%We derive a general dispersion relation for linear waves in a 
%self-gravitating, viscous, non-adiabatic disk with internal heating,
%external irradiation, and cooling. 
We analyze the {\bf axisymmetric} gravitational stability of viscous, non-adiabatic
accretion disks with internal heating, external irradiation, and
cooling {\bf in the shearing box approximation}.   
We consider parameterized cooling functions in {\bf 2D} and {\bf 3D}
disks, %two and three dimensions,  
as well as   
radiative diffusion in 3D. We show that 
generally there is no critical cooling rate/viscosity below which 
the disk is formally stable, although interesting limits appear 
for unstable modes with lengthscales on the order of the disk thickness.   
%when we examine the growth 
%of modes of order the disk scale height. 
We apply this new linear theory to protoplanetary disks subject to
gravito-turbulence modeled as an effective 
viscosity, and cooling regulated by dust opacity. 
We find that viscosity renders the disk beyond $\sim 60$AU dynamically
unstable on radial %this is for Mdot = 1d-6  
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. 
We suggest turbulent stresses can play an active role in realistic disk 
fragmentation by removing rotational stabilization against 
self-gravity, and that the observed transition in behavior from 
gravito-turbulent to fragmenting may {\bf reflect}   
instability of the gravito-turbulent state itself.
\end{abstract}

@@

{\bf 
\subsection{Instability of the gravito-turbulent state}\label{gi_of_gi}
  In this work, we formally treat the gravito-turbulent
  disk described above as an equilibrium state, to which we apply    
  standard linear stability analysis. Thus by `perturbations' we mean   
  deviations away from this gravito-turbulent 
  basic state\footnote{This should not be confused with 
    fluctuations with respect to the 
    laminar disk that maintain the underlying
    gravito-turbulence.}. 
  In our model, we interpret instability of such perturbations  to 
  signify fragmentation. Although we cannot formally demonstrate this, numerical
  simulations suggest that gravito-turbulence and fragmentations are the
  only outcomes of classic GI. Thus the failure to maintain the 
  gravito-turbulent steady state seems a reasonable description of
  fragmentation. 
%  We assume instabilities defined this way  
%  signifies disk fragmentation, since numerical simulations  
%  show that the non-existence of steady gravito-turbulence implies 
%  fragmentation.   
%Indeed, we find that the region where a gravito-turbulent PPD is
%dynamically unstable agrees with the current consensus on  where, and
%the conditions under which, a PPD is expected to fragment based on
%Eq. \ref{frag_cond}, with typical critical values cited in the
%literature

Indeed, our linear analysis for perturbations with respect to the 
gravito-turbulent state predicts that PPDs will fragment under similar
conditions observed in numerical simulations} \citep[e.g. $\beta_c\sim
  3,\,\alpha_c\sim0.1$,][]{gammie01,rice05}.

### 2016-03-28 paper/summary.tex
revised version done

More recent simulations have raised the issue of numerical
convergence. %KMK 
\cite{meru11} found that better resolved disks fragmented at longer cooling times.
Follow-up studies attributed at least some of this effect to decreasing numerical viscous
heating at higher resolution, which helps fragmentation
\citep{lodato11,meru12}. 
We can expect this if numerical viscosity
contributes to an effective irradiation, %then we do expect a disk can be
%destabilized at longer cooling times with increasing resolution, 
because then perturbations can cool to lower temperatures with
increasing resolution, see Fig. \ref{invisc_theta}.  {\bf
  In global simulations, non-convergence has also been attributed to initial 
  conditions that lead to internal edges \citep{paardekooper11b}, but
  this cannot be modeled in our local setup. 
  \cite{rice12} point out that the standard implementation of beta cooling in
  SPH simulations apply cooling on scales well-below the SPH smoothing
  lengths, and that this inconsistency may contribute to
  non-convergence. 
}

{\bf However,} \cite{paardekooper12} also found {\bf in local 2D
  grid-based simulations} that fragmentation can occur  
for  slowly-cooled disks with $\tcool\Omega\gg O(1)$, but that this requires   
simulations to run for significantly longer than dynamical 
timescales. This is consistent with our finding
that for either cooling-driven or viscous GI, there is no critical 
cooling rate/viscosity below  which the disk is  
formally stable. Instead, growth rates smoothly decrease with
increasing $\tcool$ (decreasing $\alpha$), implying that
instabilities, and hence fragmentation, simply take longer to develop
for slowly-cooled disks.         
%hopkins paper

>>>>

More recent simulations have raised the issue of numerical
convergence. %KMK 
\cite{meru11} found that better resolved disks fragmented at longer cooling times.
Follow-up studies attributed at least some of this effect to decreasing numerical viscous
heating at higher resolution, which helps fragmentation
\citep{lodato11,meru12}. 
We can expect this if numerical viscosity
contributes to an effective irradiation, %then we do expect a disk can be
%destabilized at longer cooling times with increasing resolution, 
because then perturbations can cool to lower temperatures with
increasing resolution, see Fig. \ref{invisc_theta}.  {\bf
  In global simulations, non-convergence has also been attributed to initial 
  conditions that lead to internal edges \citep{paardekooper11b}, but
  this cannot be modeled in our local setup. 
  \cite{rice12} point out that the standard implementation of beta
  cooling in smoothed-particle hydrodynamics (SPH) applies cooling on scales well-below the SPH smoothing
  lengths, and that this inconsistency may contribute to
  non-convergence. 
}

{\bf However,} \cite{paardekooper12} also found {\bf in local 2D
  grid-based simulations} that fragmentation can occur  
for  slowly-cooled disks with $\tcool\Omega\gg O(1)$, but that this requires   
simulations to run for significantly longer than dynamical 
timescales. This is consistent with our finding
that for either cooling-driven or viscous GI, there is no critical 
cooling rate/viscosity below  which the disk is  
formally stable. Instead, growth rates smoothly decrease with
increasing $\tcool$ (decreasing $\alpha$), implying that
instabilities, and hence fragmentation, simply take longer to develop
for slowly-cooled disks. 
{\bf
  However, to properly consider long timescales, it may be necessary
  to account for
  secular evolution in the global disk. 
}        
%hopkins paper
